---
title: "Customer Segmentation Analysis"
author: "Elvis Boapeah, Prince Adu Poku, Jose A Martinez Morales"
output: github_document
date: "2026-07-29"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook conducts a customer segmentation analysis on the `Purchasing_data.xlsx` data set, following the steps outlined in Part B of the assignment:

1. Select segmentation variables
2. Preprocess the data and run clustering in R
3. Visualize the segments
4. Interpret the segments and propose a marketing recommendation

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(readxl)
library(dplyr)
library(ggplot2)
library(cluster)
```
The raw file is a qualtrics export with **two header rows**: the first row holds short question codes (e.g., `Q03`, `D11`), and the second row holds the full question text. We read the codes as column names and skip the text row.

```{r load-data}
# Short codes (e.g. "Q03", "D11") become the column names
codes <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", n_max = 0) %>%
  colnames()

# Actual response data starts on row 3, so we skip the first two header rows
raw <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", skip = 2, col_names = FALSE)
colnames(raw) <- codes

dim(raw)
```
# Step 1: Exploratory Data Analysis

Before selecting and cleaning our final segmentation variables, let's take a
closer look at the data set as a whole. The goal here is simply to understand
what we're working with — not to draw conclusions yet.

We'll check:

- **Shape**. How many rows and columns the data set has    
- **Data types**.whether each column is being read as text, numeric, or
  something else, since survey data often comes in as character strings
  even when it looks numeric
- **Missing values** — which columns have gaps, and how much
- **Duplicates**Whether any respondent rows are repeated

Once we understand the shape of the data and where the gaps are, we'll be
ready to move into the actual clustering task: selecting our segmentation
variables, cleaning them, and running k-means.

```{r eda-shape}
cat("Rows:", nrow(raw), " Columns:", ncol(raw), "\n")
```
```{r eda-dtypes}
dtypes <- sapply(raw, function(x) class(x)[1])
table(dtypes)
```
```{r eda-missing}
missing_summary <- data.frame(
  Column = names(raw),
  Missing = colSums(is.na(raw)),
  Missing_Pct = round(colMeans(is.na(raw))*100, 1)
) |> arrange(desc(Missing_Pct))

head(missing_summary)
```
```{r eda-duplicates}
n_dupes <- sum(duplicated(raw))
cat("Duplicate rows:", n_dupes, "\n")
```
## What we found

- **Shape**: 4,362 rows and 89 columns, matching the data set description
  provided in the assignment.
- **Data types**: 86 columns are character (text/categorical), 1 is numeric,
  and 2 are date columns (survey start/end times tamps). This is expected for
  survey data, where most responses are recorded as category labels rather
  than raw numbers.
- **Missing values**: Missingness is heavily concentrated rather than random.
  The columns `Q13a` and `Q14` have the highest missing rate (67.8%), which
  makes sense since these questions are only shown to respondents who
  indicated they have children (`Q13`). A large block of columns is missing
  at a rate around 52–57% (e.g. `Q11_1`–`Q11_4`, `Q06`, `Q07_1`–`Q07_6`),
  consistent with survey branching and/or respondent drop-off partway
  through the questionnaire, rather than random gaps in an otherwise-complete
  response.
- **Duplicates**: No duplicate rows were found in the data set.

This tells us the data set is clean in terms of duplication, but that
completeness varies significantly by question, meaning our segmentation
variables will need to be chosen with this branching/drop-off pattern in
mind, and any missing values handled by restricting to complete cases for
the specific variables selected, rather than assuming missingness is random
across the whole data set.


## Why We're Building a Code book First

The raw data set has 89 columns, but their names (`Q01_8`, `D11`, `C1`, etc.)
are just short internal codes. They don't tell us anything on their own
about what each question actually asked. Before we can responsibly choose
segmentation variables or interpret any results, we need to understand what
every column actually represents.

Rather than looking up each column name individually as we go, we'll build
a proper code book up front: a table that maps every short code to its full
question text, and then groups related questions into categories (e.g.
Payment Methods, Credit Card Usage, Demographics). This has a few benefits:

- **It prevents mistakes.** It would be easy to confuse similarly-named
  columns (e.g. `Q07_1` through `Q07_6`, or `D02` vs. `D03`) without a clear
  reference for what each one means.
  
- **It reveals structure in the survey itself.** Grouping columns by
  category makes it clear that this survey was built from multiple
  sub-topics — payment behavior, credit card usage, BNPL attitudes, device
  ownership, and demographics — which helps explain patterns we saw earlier,
  like the block-wise missingness in our exploratory analysis.
- **It documents our reasoning for the report.** When we justify our choice
  of segmentation variables in Step 3, we can point directly to this
  code book to show why `Q03` and `D11` were selected — not just that they
  seemed reasonable, but that we understood the full landscape of
  candidate variables and considered other options (e.g. credit score,
  savings, device ownership) before making that choice.

Once the code book is built, we'll use it to confirm exactly what our chosen
segmentation variables measure, then move into cleaning and clustering.

```{r codebook-build}
labels_row <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1",
                          col_names = FALSE, skip = 1, n_max = 1)
question_text <- as.character(unlist(labels_row[1, ]))

codebook <- data.frame(Code = codes, Question = question_text, stringsAsFactors = FALSE)

head(codebook, 10)
```
```{r codebook-categories}
category_map <- list(
  "Payment Methods (12mo)"    = c("Q01_1","Q01_2","Q01_3","Q01_4","Q01_5","Q01_6","Q01_7",
                                    "Q01_8","Q01_9","Q01_10","Q01_11","Q01_12","Q01_13","Q01_14","Q01_15"),
  "Credit Card Usage"         = c("Q02","Q03","Q04","Q05","Q06"),
  "Credit Card Purchase Type" = c("Q07_1","Q07_2","Q07_3","Q07_4","Q07_5","Q07_6"),
  "BNPL Behavior"             = c("Q10","Q11_1","Q11_2","Q11_3","Q11_4"),
  "Card Choice Importance"    = c("Q12_1","Q12_2","Q12_3","Q12_4","Q12_5","Q12_6","Q12_7","Q12_8","Q12_9"),
  "Children & Credit"         = c("Q13","Q13a","Q14"),
  "Credit Score"              = c("C1"),
  "Device Ownership"          = c("D01_1","D01_4","D01_5","D01_6","D01_7","D01_8","D01_9","D01_10","D01_11","D01_12"),
  "In-Store Payment Methods"  = c("D02_1","D02_4","D02_5","D02_6","D02_7","D02_8","D02_9","D02_10","D02_11","D02_12","D02_13","D02_14"),
  "Online Payment Methods"    = c("D03_1","D03_4","D03_5","D03_6","D03_7","D03_8","D03_9","D03_10"),
  "Subscriptions"             = c("D04_1","D04_2"),
  "Demographics"              = c("D06","D07","D08","D09","D10","D11","D12","D13","D14","D15","D16"),
  "Survey Metadata"           = c("Q1","End Date")
)
```

```{r codebook-join}
code_to_category <- stack(category_map) %>% rename(Code = values, Category = ind)

codebook <- codebook %>%
  left_join(code_to_category, by = "Code") %>%
  mutate(Category = ifelse(is.na(Category), "Uncategorized", as.character(Category)))
```

```{r codebook-counts}
codebook %>% count(Category, sort = TRUE)
```
```{r codebook-full-table}
knitr::kable(codebook %>% select(Category, Code, Question), 
             caption = "Full Data Codebook")
```

# Step 3: Preparation for K-means clustering

## Preparation for the K-means clustering

Before we can run k-means clustering, our two selected variables (`Q03`
(monthly credit card balance) and `D11` (annual income)), we need to go
through several preparation steps. Right now they're stored as text
categories (e.g. `"Between $25,000 and $49,999"`), not numbers, and
k-means can't work directly with text.

Here's what we're doing, and why each step matters:

- **Trimming whitespace.** Survey exports (like this Qualtrics file)
  sometimes have stray trailing spaces on category labels. Without
  trimming, `"Two"` and `"Two "` would be treated as two *different*
  categories instead of the same one, a subtle bug that's easy to miss.

- **Converting categories to numeric midpoints.** Since both variables are
  reported as dollar *ranges* rather than exact amounts, we convert each
  range to a single representative number (its midpoint) so that k-means
  has actual numeric distances to work with. For example, `"$501 to
  $1,000"` becomes `750`. The open-ended top categories (`"More than
  $10,000"`, `"More than $100,000"`) don't have a natural midpoint, so we
  chose reasonable representative values. This is a judgment call we'll
  state explicitly as an assumption in our report.

- **Filtering to complete cases.** Not every respondent answered both
  `Q03` and `D11`. Many dropped off partway through the survey, which we
  already saw in our exploratory analysis. Rather than guessing or filling
  in missing values, we keep only the respondents who have *both* variables
  present, so that every clustering input is a real, complete observation.

- **Standardizing (scaling) the variables.** `D11` (income) ranges up into
  the hundreds of thousands, while `Q03` (balance) tops out around $12,000.
  If we clustered on the raw numbers, income would completely dominate the
  distance calculation just because its numbers are bigger, not because
  it's actually more important. Standardizing puts both variables on the
  same scale (mean 0, standard deviation 1) so they contribute equally to
  the clustering.

Once this is done, we'll have a clean, numeric, standardized dataset of
2,094 respondents, ready for k-means clustering.

````{r select-variables}
df <- raw %>%
  select(Q03, D11) %>%
  mutate(Q03 = trimws(Q03), D11 = trimws(D11))

table(df$Q03, useNA = "always")
````

````{r recode-numeric}
balance_map <- c(
  "Zero" = 0, "Less than $500" = 250, "$501 to $1,000" = 750,
  "$1,001 to $2,000" = 1500, "$2,001 to $3,000" = 2500, "$3,001 to $4,000" = 3500,
  "$4,001 to $5,000" = 4500, "$5,001 to $7,000" = 6000, "$7,001 to $10,000" = 8500,
  "More than $10,000" = 12000
)

income_map <- c(
  "Below $25,000" = 12500, "Between $25,000 and $49,999" = 37500,
  "Between $50,000 and $74,999" = 62500, "Between $75,000 and $99,999" = 87500,
  "More than $100,000" = 125000
)

df <- df %>%
  mutate(balance_num = balance_map[Q03], income_num = income_map[D11])

head(df, 10)
````

````{r filter-complete}
df_complete <- df %>% filter(!is.na(balance_num), !is.na(income_num))

cat("Complete cases:", nrow(df_complete), "of", nrow(df), "\n")
````

```{r kmeans-readiness-check}
# K-means requires numeric, complete (no NA) input
sapply(df_complete |> select(balance_num, income_num), class)
       
colSums(is.na(df_complete |> select(balance_num, income_num)))
```
```{r kmeans-readiness-summary}
summary(df_complete %>% select(balance_num, income_num))
sapply(df_complete %>% select(balance_num, income_num), sd)
```
```{r kmeans-readiness-correlation}
cor(df_complete$balance_num, df_complete$income_num)
```
```{r pre-cluster-scatter}
ggplot(df_complete, aes(x = income_num, y = balance_num)) +
  geom_jitter(alpha = 0.4, width = 3000, height = 200, color = "darkorange") +
  labs(title = "Monthly Credit Card Balance vs. Annual Income",
       x = "Annual Income (USD, midpoint)", y = "Monthly Credit Card Balance (USD, midpoint)") +
  theme_minimal()
```

## What This Plot Tells Us, And What It Doesn't

Looking at this raw scatter plot of income against credit card balance, we
can see the data is spread across a grid-like pattern (a natural result of
both variables being bucketed into ranges), with a noticeably large
concentration of respondents in the lower-balance region across every income
level, and a much smaller group of high-balance respondents scattered near
the top.

However, **it is not visually obvious from this plot alone how many distinct
customer segments actually exist** in this data, or where the boundaries
between them should be drawn. A few natural bands are visible (e.g. a dense
low-balance band, and a sparser high-balance band), but whether these should
be treated as 2 groups, 4 groups, or something else entirely isn't something
we can determine just by eye.

This is exactly the problem k-means clustering, combined with the elbow
method and silhouette score, is designed to solve: it will let the data
itself suggest a defensible number of segments (k), based on statistical
criteria rather than a visual guess.

We're now ready to move into that step: determining the appropriate value
of k, and then fitting the k-means model to formally define the customer
segments.

# Step 4: Constructing the K-Means and find the best k-value

**Standardize the variables** It is needed before clustering, since income and balance are on very different scales

```{r standardize}
X_scaled <- scale(df_complete |> select(balance_num, income_num))

head(X_scaled)
```
**Elbow method**
```{r elbow-method}
set.seed(42)
k_range <- 1:10
wss <- sapply(k_range, function(k) {
  kmeans(X_scaled, centers = k, nstart = 10)$tot.withinss
})

elbow_df <- data.frame(k = k_range, wss = wss)

ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = k_range) +
  labs(title = "Elbow Method for Optimal k",
       x = "Number of Clusters (k)", y = "Within-Cluster Sum of Squares") +
  theme_minimal()
```

**Silhouette score**

```{r silhouette-method}
set.seed(42)
sil_range <- 2:10
sil_scores <- sapply(sil_range, function(k) {
  km <- kmeans(X_scaled, centers = k, nstart = 10)
  mean(silhouette(km$cluster, dist(X_scaled))[, 3])
})

sil_df <- data.frame(k = sil_range, sil = sil_scores)

ggplot(sil_df, aes(x = k, y = sil)) +
  geom_line(color = "darkgreen") + geom_point(size = 2, color = "darkgreen") +
  scale_x_continuous(breaks = sil_range) +
  labs(title = "Silhouette Score by k",
       x = "Number of Clusters (k)", y = "Average Silhouette Width") +
  theme_minimal()

sil_df
```

**Decide on k, with reasoning documented**

```{r choose-k}
cat("Silhouette peaks at k =", sil_df$k[which.max(sil_df$sil)],
    "-- but this is inflated by ties, since both variables are coarse\n",
    "ordinal buckets (only 10 and 5 possible values). Many respondents share\n",
    "identical coordinates, letting k-means carve out many small, artificially\n",
    "'tight' clusters at high k. This is not a genuine signal of 9-10 real\n",
    "marketing segments.\n\n")

cat("WSS improvement per additional cluster:\n")
print(round(diff(wss), 1))

cat("\nThe elbow method shows WSS improvement drops sharply after k=3-4",
    "(from ~1,180 down to ~289), with only small, diminishing gains after that.",
    "k = 4 is selected: it sits at the elbow bend and produces segments that",
    "remain interpretable and actionable for marketing purposes.\n")
```