---
title: "Customer Segmentation Analysis"
author: "Elvis Boapeah, Prince Adu Poku, Jose A Martinez Morales"
output: github_document
date: "2026-07-29"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook conducts a customer segmentation analysis on the `Purchasing_data.xlsx` data set, following the steps outlined in Part B of the assignment:

1. Select segmentation variables
2. Preprocess the data and run clustering in R
3. Visualize the segments
4. Interpret the segments and propose a marketing recommendation

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(readxl)
library(dplyr)
library(ggplot2)
library(cluster)
```
The raw file is a qualtrics export with **two header rows**: the first row holds short question codes (e.g., `Q03`, `D11`), and the second row holds the full question text. We read the codes as column names and skip the text row.

```{r load-data}
# Short codes (e.g. "Q03", "D11") become the column names
codes <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", n_max = 0) %>%
  colnames()

# Actual response data starts on row 3, so we skip the first two header rows
raw <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", skip = 2, col_names = FALSE)
colnames(raw) <- codes

dim(raw)
```
# Step 2: Exploratory Data Analysis

Before selecting and cleaning our final segmentation variables, let's take a
closer look at the data set as a whole. The goal here is simply to understand
what we're working with — not to draw conclusions yet.

We'll check:

- **Shape**. How many rows and columns the data set has    
- **Data types**.whether each column is being read as text, numeric, or
  something else, since survey data often comes in as character strings
  even when it looks numeric
- **Missing values** — which columns have gaps, and how much
- **Duplicates**Whether any respondent rows are repeated

Once we understand the shape of the data and where the gaps are, we'll be
ready to move into the actual clustering task: selecting our segmentation
variables, cleaning them, and running k-means.

```{r eda-shape}
cat("Rows:", nrow(raw), " Columns:", ncol(raw), "\n")
```
```{r eda-dtypes}
dtypes <- sapply(raw, function(x) class(x)[1])
table(dtypes)
```
```{r eda-missing}
missing_summary <- data.frame(
  Column = names(raw),
  Missing = colSums(is.na(raw)),
  Missing_Pct = round(colMeans(is.na(raw))*100, 1)
) |> arrange(desc(Missing_Pct))

head(missing_summary)
```
```{r eda-duplicates}
n_dupes <- sum(duplicated(raw))
cat("Duplicate rows:", n_dupes, "\n")
```
## What we found

- **Shape**: 4,362 rows and 89 columns, matching the data set description
  provided in the assignment.
- **Data types**: 86 columns are character (text/categorical), 1 is numeric,
  and 2 are date columns (survey start/end times tamps). This is expected for
  survey data, where most responses are recorded as category labels rather
  than raw numbers.
- **Missing values**: Missingness is heavily concentrated rather than random.
  The columns `Q13a` and `Q14` have the highest missing rate (67.8%), which
  makes sense since these questions are only shown to respondents who
  indicated they have children (`Q13`). A large block of columns is missing
  at a rate around 52–57% (e.g. `Q11_1`–`Q11_4`, `Q06`, `Q07_1`–`Q07_6`),
  consistent with survey branching and/or respondent drop-off partway
  through the questionnaire, rather than random gaps in an otherwise-complete
  response.
- **Duplicates**: No duplicate rows were found in the data set.

This tells us the data set is clean in terms of duplication, but that
completeness varies significantly by question, meaning our segmentation
variables will need to be chosen with this branching/drop-off pattern in
mind, and any missing values handled by restricting to complete cases for
the specific variables selected, rather than assuming missingness is random
across the whole data set.


## Why We're Building a Code book First

The raw data set has 89 columns, but their names (`Q01_8`, `D11`, `C1`, etc.)
are just short internal codes. They don't tell us anything on their own
about what each question actually asked. Before we can responsibly choose
segmentation variables or interpret any results, we need to understand what
every column actually represents.

Rather than looking up each column name individually as we go, we'll build
a proper code book up front: a table that maps every short code to its full
question text, and then groups related questions into categories (e.g.
Payment Methods, Credit Card Usage, Demographics). This has a few benefits:

- **It prevents mistakes.** It would be easy to confuse similarly-named
  columns (e.g. `Q07_1` through `Q07_6`, or `D02` vs. `D03`) without a clear
  reference for what each one means.
  
- **It reveals structure in the survey itself.** Grouping columns by
  category makes it clear that this survey was built from multiple
  sub-topics — payment behavior, credit card usage, BNPL attitudes, device
  ownership, and demographics — which helps explain patterns we saw earlier,
  like the block-wise missingness in our exploratory analysis.
- **It documents our reasoning for the report.** When we justify our choice
  of segmentation variables in Step 3, we can point directly to this
  code book to show why `Q03` and `D11` were selected — not just that they
  seemed reasonable, but that we understood the full landscape of
  candidate variables and considered other options (e.g. credit score,
  savings, device ownership) before making that choice.

Once the code book is built, we'll use it to confirm exactly what our chosen
segmentation variables measure, then move into cleaning and clustering.

```{r codebook-build}
labels_row <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1",
                          col_names = FALSE, skip = 1, n_max = 1)
question_text <- as.character(unlist(labels_row[1, ]))

codebook <- data.frame(Code = codes, Question = question_text, stringsAsFactors = FALSE)

head(codebook, 10)
```
```{r codebook-categories}
category_map <- list(
  "Payment Methods (12mo)"    = c("Q01_1","Q01_2","Q01_3","Q01_4","Q01_5","Q01_6","Q01_7",
                                    "Q01_8","Q01_9","Q01_10","Q01_11","Q01_12","Q01_13","Q01_14","Q01_15"),
  "Credit Card Usage"         = c("Q02","Q03","Q04","Q05","Q06"),
  "Credit Card Purchase Type" = c("Q07_1","Q07_2","Q07_3","Q07_4","Q07_5","Q07_6"),
  "BNPL Behavior"             = c("Q10","Q11_1","Q11_2","Q11_3","Q11_4"),
  "Card Choice Importance"    = c("Q12_1","Q12_2","Q12_3","Q12_4","Q12_5","Q12_6","Q12_7","Q12_8","Q12_9"),
  "Children & Credit"         = c("Q13","Q13a","Q14"),
  "Credit Score"              = c("C1"),
  "Device Ownership"          = c("D01_1","D01_4","D01_5","D01_6","D01_7","D01_8","D01_9","D01_10","D01_11","D01_12"),
  "In-Store Payment Methods"  = c("D02_1","D02_4","D02_5","D02_6","D02_7","D02_8","D02_9","D02_10","D02_11","D02_12","D02_13","D02_14"),
  "Online Payment Methods"    = c("D03_1","D03_4","D03_5","D03_6","D03_7","D03_8","D03_9","D03_10"),
  "Subscriptions"             = c("D04_1","D04_2"),
  "Demographics"              = c("D06","D07","D08","D09","D10","D11","D12","D13","D14","D15","D16"),
  "Survey Metadata"           = c("Q1","End Date")
)
```

```{r codebook-join}
code_to_category <- stack(category_map) %>% rename(Code = values, Category = ind)

codebook <- codebook %>%
  left_join(code_to_category, by = "Code") %>%
  mutate(Category = ifelse(is.na(Category), "Uncategorized", as.character(Category)))
```

```{r codebook-counts}
codebook %>% count(Category, sort = TRUE)
```
```{r codebook-full-table}
knitr::kable(codebook %>% select(Category, Code, Question), 
             caption = "Full Data Codebook")
```