---
title: "Customer Segmentation Analysis"
author: "Elvis Boapeah, Prince Adu Poku, Jose A Martinez Morales"
output: github_document
date: "2026-07-29"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook conducts a customer segmentation analysis on the `Purchasing_data.xlsx` data set, following the steps outlined in Part B of the assignment:

1. Select segmentation variables
2. Preprocess the data and run clustering in R
3. Visualize the segments
4. Interpret the segments and propose a marketing recommendation

We'll build this up step by step, one chunk at a time.

# Step 0: Load Packages and Data

```{r load-packages}
library(readxl)
library(dplyr)
library(ggplot2)
library(cluster)
```
The raw file is a qualtrics export with **two header rows**: the first row holds short question codes (e.g., `Q03`, `D11`), and the second row holds the full question text. We read the codes as column names and skip the text row.

```{r load-data}
# Short codes (e.g. "Q03", "D11") become the column names
codes <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", n_max = 0) %>%
  colnames()

# Actual response data starts on row 3, so we skip the first two header rows
raw <- read_excel("data/Purchasing_data.xlsx", sheet = "Sheet1", skip = 2, col_names = FALSE)
colnames(raw) <- codes

dim(raw)
```
# Step 2: Exploratory Data Analysis

Before selecting and cleaning our final segmentation variables, let's take a
closer look at the data set as a whole. The goal here is simply to understand
what we're working with — not to draw conclusions yet.

We'll check:

- **Shape**. How many rows and columns the data set has    
- **Data types**.whether each column is being read as text, numeric, or
  something else, since survey data often comes in as character strings
  even when it looks numeric
- **Missing values** — which columns have gaps, and how much
- **Duplicates**Whether any respondent rows are repeated

Once we understand the shape of the data and where the gaps are, we'll be
ready to move into the actual clustering task: selecting our segmentation
variables, cleaning them, and running k-means.

```{r eda-shape}
cat("Rows:", nrow(raw), " Columns:", ncol(raw), "\n")
```
```{r eda-dtypes}
dtypes <- sapply(raw, function(x) class(x)[1])
table(dtypes)
```
```{r eda-missing}
missing_summary <- data.frame(
  Column = names(raw),
  Missing = colSums(is.na(raw)),
  Missing_Pct = round(colMeans(is.na(raw))*100, 1)
) |> arrange(desc(Missing_Pct))

head(missing_summary)
```
```{r eda-duplicates}
n_dupes <- sum(duplicated(raw))
cat("Duplicate rows:", n_dupes, "\n")
```
## What we found

- **Shape**: 4,362 rows and 89 columns, matching the data set description
  provided in the assignment.
- **Data types**: 86 columns are character (text/categorical), 1 is numeric,
  and 2 are date columns (survey start/end times tamps). This is expected for
  survey data, where most responses are recorded as category labels rather
  than raw numbers.
- **Missing values**: Missingness is heavily concentrated rather than random.
  The columns `Q13a` and `Q14` have the highest missing rate (67.8%), which
  makes sense since these questions are only shown to respondents who
  indicated they have children (`Q13`). A large block of columns is missing
  at a rate around 52–57% (e.g. `Q11_1`–`Q11_4`, `Q06`, `Q07_1`–`Q07_6`),
  consistent with survey branching and/or respondent drop-off partway
  through the questionnaire, rather than random gaps in an otherwise-complete
  response.
- **Duplicates**: No duplicate rows were found in the data set.

This tells us the data set is clean in terms of duplication, but that
completeness varies significantly by question, meaning our segmentation
variables will need to be chosen with this branching/drop-off pattern in
mind, and any missing values handled by restricting to complete cases for
the specific variables selected, rather than assuming missingness is random
across the whole data set.
